Exact Asymptotics for Causal Mediation Analysis

### 2024-05-24 notes/William Analysis.Rmd
Run analysis for bin-bin-mix case

```{r}
# Values of X and W for which we compute the total effect
x_pred = 0
W_pred = c(1,1,1)

# Values of a_0 and b_0
a_0 = 0
b_0 = -0.5
```

```{r MC-bin-bin-fix, include=FALSE, cache=TRUE}
# , dependson=c("boilerplate")}

    ## Generate Y
    zeta_vec = b_0 + b_1*M + b_2*X + W%*%B_3
    p_Y_vec = expit(zeta_vec)
    Y = rbinom(n, size = 1, prob = p_Y_vec)

    ## Y
    Y_data = data.frame(Y, M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
    Y_model = glm(Y ~ M + X + W1 + W2 + W3, data = Y_data, family = binomial(link = "logit"))

    ## Linear predictor for M
    eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat
    zeta_hat = b_0_hat + b_x_hat*x_pred + W_pred %*% B_3_hat

    # Mediation effect
    ## See Helpers.R for the function get_odds_ratio
    med_hat = get_odds_ratio(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat)
    some_med_hats[[i]] = med_hat

    ## Compute gradient of mediation effect
    grad_med = d_OR_d_theta(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat, x_pred, W_pred)

@@

a_hat = fixef(M_model)
a_RE_sds = M_model_info$stddev
a_RE_cor = M_model_info$correlation[2,1]
a_cov = vcov(M_model, full=TRUE, ranpar="sd")

@@

b_hat = fixef(Y_model)
b_RE_sds = Y_model_info$stddev
b_RE_cor = Y_model_info$correlation[2,1]
b_cov = vcov(Y_model, full=TRUE, ranpar="sd")

@@

@@

@@

@@

@@

>>>>

@@

a_hat = fixef(M_model)
a_RE_sds = M_model_info$stddev
a_RE_cor = M_model_info$correlation[2,1]
theta_hat = c(a_RE_sds, a_RE_cor)
M_cov = vcov(M_model, full=TRUE, ranpar="sd")

@@

b_hat = fixef(Y_model)
b_RE_sds = Y_model_info$stddev
b_RE_cor = Y_model_info$correlation[2,1]
gamma_hat = c(b_RE_sds, b_RE_cor)
Y_cov = vcov(Y_model, full=TRUE, ranpar="sd")

@@

```{r}

# Fixed-effects
x_pred = 0
W_pred = c(1,1,1)

@@

# Linear predictors
eta_hat = as.numeric(a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat)
zeta_hat = as.numeric(b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat)

# Random effects covariances
s_M_0 = a_RE_sds[1]
s_M_x = a_RE_sds[2]
rho_M = a_RE_cor

s_Y_0 = b_RE_sds[1]
s_Y_x = b_RE_sds[2]
rho_Y = b_RE_cor

# Sigma functions
sigma_M1 = sigma_fun(x_pred, s_M_0, s_M_x, rho_M)
sigma_M2 = sigma_fun(x_pred + 1, s_M_x, s_M_0, rho_M)

sigma_Y1 = sigma_fun(x_pred, s_Y_0, s_Y_x, rho_Y)
sigma_Y2 = sigma_fun(x_pred + 1, s_Y_x, s_Y_0, rho_Y)

# Mediation effect
## See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
```

Applying the $\delta$-method here is a bit more complicated than in the other settings considered so far. Following the notes written by B & B, we take a two-step approach. We first transform model parameters from the two GLMMs into the vector of arguments to $\Phi$; call such a function $\xi$. We then compute the gradient of $\Phi$ with respect to these arguments, and use the $\delta$-method to get the standard error.

In order to perform the first step, we need the gradient of $\xi$. Fortunately, much of this has already been done for us. Specifically, B & B worked out the gradients of each of the four $\sigma$ functions. All that remains is to repeat this process for the other arguments of $\Phi$, and stack the results in a matrix. See the `Helpers.R` file for the relevant functions.

Starting first with 
We now compute the gradient of the odds ratio and use the $\delta$-method to get the standard error.

```{r}
# Gradient of OR wrt regression coefficients
grad_Phi_obs = grad_Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
grad_xi_obs = grad_xi(a_hat, theta_hat, b_hat, gamma_hat, x_pred, W_pred)
d_Phi_d_GLMM_pars = grad_Phi_obs %*% grad_xi_obs

@@

## Build joint covariance matrix of regression coefficients
M_length = nrow(M_cov)
Y_length = nrow(Y_cov)
joint_cov = matrix(0, nrow = M_length + Y_length, ncol = M_length +
Y_length)
joint_cov[1:M_length, 1:M_length] = M_cov
joint_cov[(M_length + 1):(M_length + Y_length), (M_length + 1):(M_length +
Y_length)] = Y_cov

@@

## Pre- and post-multiply asymptotic covariance by gradient of Phi wrt GLMM parameters
med_asymp_var = d_Phi_d_GLMM_pars %*% asymp_reg_cov %*% t(d_Phi_d_GLMM_pars)

@@

```{r, include=FALSE}
<<boilerplate>>

library(lme4)     # For fitting GLMMs using glmer()
library(MASS)     # For simulating multivariate normals using mvrnorm()
library(merDeriv) # For computing the full information matrix from models fit using lme4

# Fewer large sample sizes
all_Ns = c(100, 200, 500, 1000, 2000)

# Values of X and W for which we compute the total effect
x_pred = 0
W_pred = c(1,1,1)

# Fixed effects for intercepts
a_0 = 0
a = c(a_0, a_1, A_2)
b_0 = -0.5
b = c(b_0, b_1, b_2, B_3)

# Number of clusters
K=5

# Number of replicates for each sample size
# We use fewer here because model fitting takes longer
num_reps = 200
```

```{r MC-bin-bin-mix, include=FALSE, cache=TRUE}
# , dependson=c("boilerplate")}

tic()

# Containers for output
all_a_hats = list()
all_theta_hats = list()
all_b_hats = list()
all_gamma_hats = list()

all_M_covs = list()
all_Y_covs = list()

  # Containers for output with this value of n
  some_a_hats = list()
  some_theta_hats = list()
  some_b_hats = list()
  some_gamma_hats = list()

  some_M_covs = list()
  some_Y_covs = list()

  for(i in 1:num_reps){
    print(paste0("n = ", n, ", i = ", i, " of ", num_reps))

    # Use tryCatch to omit any datasets leading to convergence issues
    tryCatch({
      all_Xs = list()
      all_Ws = list()

      # Consolidate groups
      X = do.call(c, all_Xs)
      W = do.call(rbind, all_Ws)
      M = do.call(c, all_Ms)
      Y = do.call(c, all_Ys)
      group = rep(1:K, each = n)

      ## M
      M_data = data.frame(M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
      M_model = glmer(M ~ X + W1 + W2 + W3 + (1 + X | group), data = M_data, family = binomial(link = "logit"))
      M_model_info = attributes(VarCorr(M_model)$group)

      ### Fitted parameters
      a_hat = fixef(M_model)
      a_RE_sds = M_model_info$stddev
      a_RE_cor = M_model_info$correlation[2,1]
      theta_hat = c(a_RE_sds, a_RE_cor)
      if(any(is.nan(theta_hat))) stop("NaNs in theta_hat")  # Skip rest of current analysis if correlation is 0/0

      ### Estimated SE covariance matrix
      M_cov = vcov(M_model, full=TRUE, ranpar="sd")

      some_a_hats[[i]] = a_hat
      some_theta_hats[[i]] = theta_hat
      some_M_covs[[i]] = M_cov

      ## Y
      Y_data = data.frame(Y, M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
      Y_model = glmer(Y ~ M + X + W1 + W2 + W3 + (1 + X | group), data = Y_data, family = binomial(link = "logit"))
      Y_model_info = attributes(VarCorr(Y_model)$group)

      ### Fitted parameters
      b_hat = fixef(Y_model)
      b_RE_sds = Y_model_info$stddev
      b_RE_cor = Y_model_info$correlation[2,1]
      gamma_hat = c(b_RE_sds, b_RE_cor)
      if(any(is.nan(gamma_hat))) stop("NaNs in gamma_hat")  # Skip rest of current analysis if correlation is 0/0

      ### Estimated SE covariance matrix
      Y_cov = vcov(Y_model, full=TRUE, ranpar="sd")

      some_b_hats[[i]] = b_hat
      some_gamma_hats[[i]] = gamma_hat
      some_Y_covs[[i]] = Y_cov

      ## Fixed-effects
      x_pred = 0
      W_pred = c(1,1,1)

      ## Linear predictors
      eta_hat = as.numeric(a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat)
      zeta_hat = as.numeric(b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat)

      ## Random effects covariances
      s_M_0 = a_RE_sds[1]
      s_M_x = a_RE_sds[2]
      rho_M = a_RE_cor

      s_Y_0 = b_RE_sds[1]
      s_Y_x = b_RE_sds[2]
      rho_Y = b_RE_cor

      ## Sigma functions
      sigma_M1 = sigma_fun(x_pred, s_M_0, s_M_x, rho_M)
      sigma_M2 = sigma_fun(x_pred + 1, s_M_x, s_M_0, rho_M)

      sigma_Y1 = sigma_fun(x_pred, s_Y_0, s_Y_x, rho_Y)
      sigma_Y2 = sigma_fun(x_pred + 1, s_Y_x, s_Y_0, rho_Y)

      ## Mediation effect
      ### See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
      med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
      some_med_hats[[i]] = med_hat

      ## Gradient of OR wrt regression coefficients
      grad_Phi_obs = grad_Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
      grad_xi_obs = grad_xi(a_hat, theta_hat, b_hat, gamma_hat, x_pred, W_pred)
      d_Phi_d_GLMM_pars = grad_Phi_obs %*% grad_xi_obs

      ## Get asymptotic SE using delta method

      ### Build joint covariance matrix of regression coefficients
      M_length = nrow(M_cov)
      Y_length = nrow(Y_cov)
      joint_cov = matrix(0, nrow = M_length + Y_length, ncol = M_length +
      Y_length)
      joint_cov[1:M_length, 1:M_length] = M_cov
      joint_cov[(M_length + 1):(M_length + Y_length), (M_length + 1):(M_length +
      Y_length)] = Y_cov

      ### Convert to asymptotic covariance matrix
      asymp_reg_cov = n * joint_cov

      ### Pre- and post-multiply asymptotic covariance by gradient of Phi wrt GLMM parameters
      med_asymp_var = d_Phi_d_GLMM_pars %*% asymp_reg_cov %*% t(d_Phi_d_GLMM_pars)

      ### Get small-sample standard error
      med_asymp_SE = sqrt(med_asymp_var)
      med_SE = med_asymp_SE / sqrt(n)

      some_med_SEs[[i]] = med_SE
    }, error = function(e){})
  }

  # Store output for current value of n
  all_a_hats[[j]] = some_a_hats
  all_theta_hats[[j]] = some_theta_hats
  all_b_hats[[j]] = some_b_hats
  all_gamma_hats[[j]] = some_gamma_hats

  all_M_covs[[j]] = some_M_covs
  all_Y_covs[[j]] = some_Y_covs

  all_med_hats[[j]] = some_med_hats
  all_med_SEs[[j]] = some_med_SEs
}
toc()

```{r}
# Re-format output to be more useful. 
data_med_hats = lapply(all_med_hats, unlist)
data_med_SEs = all_med_SEs %>% lapply(unlist) %>% lapply(na.omit)

# Combine into a table
results = data.frame(Empirical = emp_SEs, 
                     Mean = mean_delta_SEs, Median = median_delta_SEs,
                     "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
                     "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
# results = data.frame(n = all_Ns, Empirical = emp_SEs, 
#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)

I don't really understand the results I'm getting here. It should help if I can compare my estimates of the mediation effect with the true value.

```{r}
